Pillar 1 - Sequential Pattern Mining

### 2019-02-14 Pillar 1 - Sequential Pattern Mining.Rmd
Major re-organization of script: functions to make rules before making rules

# Setup
We load the environment `data_pillar1` which we saved at the end of the data preparation. It contains the data sets `d_course` and `d_transcript`.
```{r loading data}
load("Output/data_pillar_1.RDATA")
```
We set the thresholds for passing and obtaining a high grade.
```{r}
pass_grade <- 5.5
high_grade <- 6.5
```

@@

```{r statistic summary}

@@

# for convenience
join_d_course <- function(data){

@@

# Concentration evel
statistics$concentration <- d_transcript %>%
  join_d_course %>%
  gather(
    key = X, 
    value = Concentration, 
    Concentration, `Concentration (additional)`,
    na.rm = TRUE
    ) %>%
  provide_statistics(Concentration)

@@

# Level level
statistics$level <- d_transcript %>%
  # TODO: filter for student who completed their studies
  join_d_course %>%
  provide_statistics(Level)

#
# output
save(
  statistics,
  file = "Output/Transcript Statistics.RDATA"
  )

rm(statistics)
```

# Creating Transactions and Sequences
For a first exploration of arules, we conceptualise our framework like this:
sequence = student
item     = course

## Data Preparation
First we transform our data into transaction data. For this, we first create a vector of mandatory courses that we exclude from transcripts. Furthermore, for courses that span more than one period, we consider only the starting point.

### Take, fail, low grade
For our first transactions we add variables that code whether a course has been passed/failed, or passed with a high grade/low grade. We assign the `Student ID` to the variable `sequenceID`, and the timing of a course (Year,period) to the `eventID`. We identify three different sets of items (course, course_pass/course_fail, course_high/course_low) for the creation of different rules further on. 
```{r AR data prep 1}
d_transactions <- list()

@@

  # Exclude 
  filter(
    Type != "Mandatory",                  # (i) mandatory courses e.g. COR, CAP, etc
    ! Letters %in% c("SKI", "PRO",        # (ii) skills & projects (taken by majority of students)
                     "SAS", "SAH", "SAC") # (iii) courses of semester abroad (uninformative)
    ) %>%

  # Join with transcripts.
  select(- Period) %>%
  inner_join(
    d_transcript,
    by = "Course ID"
    ) %>%

  # Identifying sequenceID
  rename(sequenceID = `Student ID`) %>%

@@

  # Identifying itemID
  mutate(
    PF = case_when( Grade <  pass_grade ~ "fail",
                    Grade >= pass_grade ~ "pass"),
    HL = case_when( Grade <  high_grade ~ "low",
                    Grade >= high_grade ~ "high"),

    item    = `Course ID`,
    item_PF = paste(`Course ID`, PF, sep = "_"),
    item_HL = paste(`Course ID`, HL, sep = "_")
    )

```

### Take / not take, pass / fail / not take
For our second set of transactions, we come up with all possible combinations of courses and students, then categorise each as, not taken, passed or failed. Items are the courses with their pass/fail/not taken categories (course_pass/course_fail/course_not taken).
```{r transactions not taken}
d_transactions$TPF <- expand.grid(

  # Expand along students (sequenceID) and courses (itemID)
  sequenceID = unique(d_transactions$taken_PF_HL$sequenceID),
  item       = unique(d_transactions$taken_PF_HL$item),
  stringsAsFactors = FALSE
  ) %>%

@@

  # Create 
  mutate(

    TPF = case_when(
      is.na(Grade)       ~ "not taken",
      Grade < pass_grade ~ "fail",
      TRUE               ~ "pass"
      ),

@@

rm(high_grade, pass_grade)
```

### Grade
For our last set of transactions we would like to have cumulative probabilities so we can interpret a prediction of a grade 8 as the probability of getting an 8 or lower. 
For this, we add additional rows to the transcripts, corresponding to all the grades above the grade that a student got for a course (e.g. if student_A got a grade_8 for course_B, we would add two rows: student_A/course_B/grade_9 and student_A/course_B/grade_10). Items in these transactions correspond to course_grade.
```{r transactions grade}

d_transactions$G <- d_transactions$taken_PF_HL %>%

  # Spread along rounded grades
  mutate(
    grade_ceil = ceiling(Grade),
    Values     = TRUE
    ) %>%
  spread(
    key   = grade_ceil, 
    value = Values
    ) %>%

  # fill grades inferior to obtained grade with TRUE
  mutate(
    `1`  = `0`|`1`,
    `2`  = `1`|`2`,
    `3`  = `2`|`3`,
    `4`  = `3`|`4`,
    `5`  = `4`|`5`,
    `6`  = `5`|`6`,
    `7`  = `6`|`7`,
    `8`  = `7`|`8`,
    `9`  = `8`|`9`,
    `10` = `9`|`10`
    ) %>%

@@

### Transcript with preceding courses (for SR by hand)
For our analysis in sequential rule mining, we extract the history of each student, and create a transcript per student, per period, which contains information on the current courses, the courses taken before the current courses, courses taken so far (including current courses) and the courses not taken so far. 
```{r d_transcript_cum}
#
# List of all courses
course_all <- unique(d_transactions$taken_PF_HL$item)

#
# courses taken so far and not taken so far
d_transcript_cum <- d_transactions$taken_PF_HL %>%

@@

  # course current
  group_by(
    sequenceID,
    eventID
    ) %>%
  summarize_at(
    c("item", "item_take", "item_PF", "item_HL"),
    function(x) list(unique(x))
    ) %>%
  ungroup %>%

  # course current all
  rowwise %>%
  mutate(
    course_current = list(c(item, item_take, item_PF, item_HL))
    ) %>%
  ungroup %>%

  # course past
  group_by(
    sequenceID
    ) %>%
  mutate(
    course_past = Reduce(
      f = c,
      x = course_current,
      accumulate = TRUE
      ) %>%
      lag(
        default = as.character(NA)
        )
    ) %>%
  ungroup %>%

  # course so far, course not so far
  rowwise %>%
  mutate(
    course_past_current = list(union  (course_past, course_current      )),
    course_not_yet      = list(setdiff(course_all , course_past_current )),
    course_not_yet      = list(paste(course_not_yet, "not", sep = "_"   )),
    course_so_far       = list(union  (course_past, course_not_yet))
    ) %>%
  ungroup %>%

  # lhs: past, rhs: present
  select(
    sequenceID,
    lhs = course_so_far,
    rhs = course_current
  )

rm(course_all)
```

## Transactions
The function make_transactions takes a data.frame and the item label and converts it into an object of class transaction. 
We create transactions with the different types of data we prepared above (i.e. taken, pass/fail, high/low, NotTaken/pass/fail, Grades).
```{r making transactions, message = FALSE}

make_transaction <- function(data = d_transactions$taken_PF_HL, item = item){

@@

#
# Checking transactions
# inspect(head(transactions, 10))

rm(make_transaction)
```

## Sequences
The function make_sequence takes a data.frame and the item label and converts it into an object of class sequences (used for extracting sequential rules).
We create sequences with the different types of data we prepared above (i.e. taken, pass/fail, high/low, Grades).
```{r making sequences}

make_sequence <- function(data = d_transactions$taken_PF_HL, item = item){

    # make transactions
  data_temp <- data %>%
    arrange(
      sequenceID,
      eventID
    ) %>%
    filter(
      !is.na(eventID) #this is here casue we have courses that were never taken.
      ) %>% 
    group_by(
      sequenceID,
      eventID
    ) %>%
    summarise(
      list_item = list(unique(!!enquo(item)))
    ) %>%
    ungroup

@@

  # indicating sequence ID and event ID for each transaction
  sequences@itemsetInfo <- data_temp %>%
    select(
      sequenceID,
      eventID
      ) %>%
    as.data.frame

@@

sequences$taken <- make_sequence(item = item   )
sequences$PF    <- make_sequence(item = item_PF)
sequences$HL    <- make_sequence(item = item_HL)
#sequences_TPF  <- make_sequence(data = d_transactions$TPF, item = item_TPF) # compute ad hoc
sequences$G     <- make_sequence(data = d_transactions$G    , item = item_G  )

#
# Checking sequences
 inspect(head(sequences$G, 10))

rm(make_sequence #, d_transactions
   )
```

## Support
```{r}

#
# Number of students
n_students <- d_transactions$taken_PF_HL %>%
  select(sequenceID) %>%
  n_distinct

#
# Support
d_support <- list()

#
# Probability of taking a course
d_support$TNT <- d_transactions$taken_PF_HL %>%
  distinct(
    item,
    sequenceID
    ) %>%
  count(
    item
    ) %>%
  mutate(
    rate.take = n / n_students,
    rate.not  = 1 - rate.take
    )

@@

# Mining Rules
## Association Rule

### Functions

@@

The function `clean_AR()` transforms the rules generated by the function `my_apriori()` into a readable dataframe

```{r function clean_rules}
clean_rules <- function(rules){

The function `compute_support()` computes the support of the lhs and rhs of the rules.
```{r compute_support}
compute_rhs.support <- function(AR, data_support, type_rule){

}
```
###Apply Functions
We apply make_AR to get different sets of rules:

#### take -> take
```{r AR take}
AR <- list()

#### fail -> fail
```{r AR fail}
AR$PF <- transactions$PF %>%

#### low grade -> low grade
```{r AR low}
AR$HL <- transactions$HL %>%

#### not take -> fail
```{r AR TPF}
AR$TPF <- transactions$TPF %>%

#### grade less than or equal to x -> grade less than or equal to 6
```{r AR grade, cache = TRUE}
AR$G <- transactions$G %>%

  my_apriori %>%
  clean_rules %>%

### Editing AR
The function edit_rules makes AR easier to read. It keeps only rules that appear more than 5 times, rounds numerical variables to 5 significant digits, and drops aiding columns which were only used for computation in previous stages but add no additional information.
```{r edit_AR}
edit_rules <- function(AR){

AR <- lapply(
  X   = AR,
  FUN = edit_rules
  )
```

### Save Association Rules
We save and remove objects:
```{r saving and rm}
#
# Save association rules
save(
  AR,
  file = "App/AR.RDATA"
  )

# Remove objects
rm(
  my_apriori, make_AR,
  AR, transaction
  )

```

## Sequence Rules
### Functions
The function `my_cspade()` applies the cspade algorithm on a set of sequential transactions with the parameters that we have chosen.
```{r my_cspade function}
my_cspade <- function(data){

@@

}
```
The function `my_ruleInduction()` creates rules from the set of frequent sequences detemined by cspade.
```{r my_ruleInduction}
my_ruleInduction <- function(rules){

@@

}
```
We encapsulate the fuctions `my_cpade`, `my_ruleInduction`, and `clean_rules` into `make_SR`
```{r make_SR}
make_SR <- function(data, data_support, type_rule){

@@

}
```
###Apply Functions
We apply make_SR to get different types of sequential rules:
#### take -> take
```{r SR taken}
SR <- list()

@@

#### fail -> fail
```{r SR PF}
SR$PF <- sequences$PF %>%

@@

#### grade less than or equal to x -> grade less than or equal to 6
```{r SR G, cache = TRUE}
SR$G <- sequences$G %>%

@@

### Editing SR
We apply `edit_rules` to SR in order to get easily readable rules. 
```{r SR edit}
SR <- lapply(
  X   = SR,
  FUN = edit_rules
  )
```

### Save Sequences Rules
```{r SR save}
#
# Save Sequence rules
save(
  SR,
  file = "App/SR.RDATA"
)

#
# Remove objects
rm(
  SR, sequences,
  my_cspade, my_ruleInduction, clean_rules, compute_rhs.support, edit_rules, make_SR,
  n_students
)
```

>>>>

# Setup
We load the environment `data_pillar1` which we saved at the end of the data preparation. It contains the data sets `d_course` and `d_transcript`.
```{r loading data}
load("Output/data_pillar_1.RDATA")
```

```{r for convenience}
pass_grade <- 5.5
high_grade <- 6.5

join_d_course <- function(data){

join_d_transcript <- function(data){

  data %>%
    inner_join(
    d_transcript,
    by = c("Course ID")
    )

@@

```{r statistic summary, cache = TRUE}

@@

@@

# Concentration level
statistics$concentration <- d_transcript %>%
  join_d_course %>%
  gather(
    key = X, 
    value = Concentration, 
    Concentration, `Concentration (additional)`,
    na.rm = TRUE
    ) %>%
  provide_statistics(Concentration)

@@

# Level level
statistics$level <- d_transcript %>%
  # TODO: filter for student who completed their studies
  join_d_course %>%
  provide_statistics(Level)
```

```{r, include = FALSE}
save(
  statistics,
  file = "Output/Transcript Statistics.RDATA"
  )
```

```{r, include = FALSE}
rm(statistics, provide_statistics)
```

# Creating Transactions and Sequences
For a first exploration of arules, we conceptualise our framework like this:
transaction = student
item = course

First we transform our data into transaction data. For this, we first create a vector of mandatory courses that we exclude from transcripts. 
## Data Preparation

### Take, fail, low grade
```{r d_transaction taken_PF_HL}
d_transactions <- list()

@@

  # Exclude 
  filter(
    Type != "Mandatory",  # (i) mandatory courses e.g. COR, CAP, etc
    ! Letters %in% c(
      "SKI", "PRO",       # (ii) skills & projects (taken by majority of students)
      "SAS", "SAH", "SAC" # (iii) courses of semester abroad (uninformative)
      )
    ) %>%

  # Join with transcripts
  select(
    - Period
    ) %>%
  join_d_transcript %>%

  # Identifying sequenceID
  rename(
    sequenceID = `Student ID`
    ) %>%

@@

  # Identifying itemID
  mutate(
    PF = case_when(
      Grade <  pass_grade ~ "fail",
      Grade >= pass_grade ~ "pass"
      ),
    HL = case_when(
      Grade <  high_grade ~ "low",
      Grade >= high_grade ~ "high"
      ),

    item    = `Course ID`,
    item_PF = paste(item, PF, sep = "_"),
    item_HL = paste(item, HL, sep = "_")
    )
```

```{r for convenience II}
course_all  <- unique(d_transactions$taken_PF_HL$item      )
student_all <- unique(d_transactions$taken_PF_HL$sequenceID)
n_students  <- length(student_all)
```

### Take / not take, pass / fail / not take
```{r d_transactions TPF}
d_transactions$TPF <- expand.grid(

  # Expand along students (sequenceID) and courses (itemID)
  sequenceID       = student_all,
  item             = course_all,
  stringsAsFactors = FALSE
  ) %>%

@@

  # Create item
  mutate(

    TPF = case_when(
      is.na(Grade)        ~ "not taken",
      Grade <  pass_grade ~ "fail",
      Grade >= pass_grade ~ "pass"
      ),

@@

    )
```

### Grade
```{r transactions grade}
d_transactions$G <- d_transactions$taken_PF_HL %>%

  # Spread along ceiling grades
  mutate(
    grade_ceil = ceiling(Grade),
    Values     = TRUE
    ) %>%
  spread(
    key   = grade_ceil, 
    value = Values
    ) %>%

  # grade x or lower
  mutate(
    `1`  = `0`|`1`,
    `2`  = `1`|`2`,
    `3`  = `2`|`3`,
    `4`  = `3`|`4`,
    `5`  = `4`|`5`,
    `6`  = `5`|`6`,
    `7`  = `6`|`7`,
    `8`  = `7`|`8`,
    `9`  = `8`|`9`,
    `10` = `9`|`10`
    ) %>%

@@

### Transcript with preceding courses (for SR by hand)
```{r d_transcript_cum, cache = TRUE, warning = FALSE}
d_transcript_cum <- d_transactions$taken_PF_HL %>%

@@

  # course current
  group_by(
    sequenceID,
    eventID
    ) %>%
  summarize_at(
    c("item", "item_take", "item_PF", "item_HL"),
    function(x) list(unique(x))
    ) %>%

  # course current all
  rowwise %>%
  mutate(
    course_current = list(c(item, item_take, item_PF, item_HL))
    ) %>%

  # course past
  group_by(
    sequenceID
    ) %>%
  mutate(
    course_past = Reduce(
      f = c,
      x = course_current,
      accumulate = TRUE
      ) %>%
      lag(
        default = as.character(NA)
        )
    ) %>%

  # course so far, course not so far
  rowwise %>%
  mutate(

    course_past_current = list(union  (course_past, course_current      )),

    course_not_yet      = list(setdiff(course_all , course_past_current )),
    course_not_yet      = list(paste(course_not_yet, "not", sep = "_"   )),

    course_so_far       = list(union  (course_past, course_not_yet))

    ) %>%
  ungroup %>%

  # lhs: past, rhs: present
  select(
    sequenceID,
    lhs = course_so_far,
    rhs = course_current
  )
```

## Transactions
```{r making transactions, cache = TRUE}
make_transaction <- function(data = d_transactions$taken_PF_HL, item = item){

@@

# inspect(head(transactions, 10))
```

## Sequences
```{r making sequences, cache = TRUE}
make_sequence <- function(data = d_transactions$taken_PF_HL, item = item){

  data_temp <- data %>%
    arrange(
      sequenceID,
      eventID
    ) %>%
    filter(
      !is.na(eventID) # exclude courses that were never taken.
      ) %>% 
    group_by(
      sequenceID,
      eventID
    ) %>%
    summarise(
      list_item = list(unique(!!enquo(item)))
    ) %>%
    ungroup

@@

  # indicate sequence ID and event ID for each transaction
  sequences@itemsetInfo <- data_temp %>%
    select(
      sequenceID,
      eventID
      ) %>%
    as.data.frame

@@

sequences$taken <- make_sequence(item = item   )
sequences$PF    <- make_sequence(item = item_PF)
sequences$HL    <- make_sequence(item = item_HL)
sequences$G     <- make_sequence(data = d_transactions$G    , item = item_G  )

#inspect(head(sequences$G, 10))
```

## Support
```{r}
d_support <- list()

# Probability of taking a course
d_support$TNT <- d_transactions$taken_PF_HL %>%
  distinct(
    item,
    sequenceID
    ) %>%
  count(
    item
    ) %>%
  mutate(
    rate.take = n / n_students,
    rate.not  = 1 - rate.take
    )

@@

```{r, include = FALSE}
save(
  transactions, sequences,
  d_transcript_cum,
  d_support, n_students,
  file = "Output/Transaction and Sequences.RDATA"
  )
```

```{r, include = FALSE}
rm(

  high_grade, pass_grade, 
  course_all, student_all, n_students,

  d_transactions, d_support, d_transcript_cum,

  make_transaction, transactions, 
  make_sequence, sequences
  )
```

# Mining Rules

```{r}
load("Output/Transaction and Sequences.RDATA")
```

## Functions

### my_aprior, my_cspade and my_ruleInduction

@@

The function `my_cspade()` applies the cspade algorithm on a set of sequential transactions with the parameters that we have chosen.
```{r my_cspade function}
my_cspade <- function(data){

@@

The function `my_ruleInduction()` creates rules from the set of frequent sequences detemined by cspade.
```{r my_ruleInduction}
my_ruleInduction <- function(rules){

@@

### clean_rules
The function `clean_AR()` transforms the rules generated by the function `my_apriori()` into a readable dataframe
```{r function clean_rules}
clean_rules <- function(rules){

### compute_rhs.support
The function `compute_support()` computes the support of the lhs and rhs of the rules.
```{r compute_support}
compute_rhs.support <- function(AR, data_support, type_rule){

### make_AR and  make_SR (wrapers)

We encapsulate the fuctions `my_cpade`, `my_ruleInduction`, and `clean_rules` into `make_SR`
```{r make_SR}
make_SR <- function(data, data_support, type_rule){

@@

## Association Rule
We apply make_AR to get different sets of rules:

### take -> take
```{r AR take, cache = TRUE, warning = FALSE}
AR <- list()

### fail -> fail
```{r AR fail, cache = TRUE}
AR$PF <- transactions$PF %>%

### low grade -> low grade
```{r AR low, cache = TRUE}
AR$HL <- transactions$HL %>%

### not take -> fail
```{r AR TPF, cache = TRUE}
AR$TPF <- transactions$TPF %>%

### grade less than or equal to x -> grade less than or equal to 6
```{r AR grade, cache = TRUE}
AR$G <- transactions$G %>%

  my_apriori %>%

  clean_rules %>%

## Sequence Rules
We apply make_SR to get different types of sequential rules:

### take -> take
```{r SR taken, cache = TRUE}
SR <- list()

@@

### fail -> fail
```{r SR PF, cache = TRUE}
SR$PF <- sequences$PF %>%

@@

### grade less than or equal to x -> grade less than or equal to 6
```{r SR G, cache = TRUE}
SR$G <- sequences$G %>%

@@

## Editing rules
The function edit_rules makes AR easier to read. It keeps only rules that appear more than 5 times, rounds numerical variables to 5 significant digits, and drops aiding columns which were only used for computation in previous stages but add no additional information.
```{r edit_rules}
edit_rules <- function(AR){

```{r edit rules}
AR <- lapply(
  X   = AR,
  FUN = edit_rules
  )

SR <- lapply(
  X   = SR,
  FUN = edit_rules
  )
```

```{r show rules}
print(AR$TPF)
print(SR$HL)
```

```{r save rules, include = FALSE}
save(
  AR, SR,
  file = "App/rules.RDATA"
  )
```
